---
title: "Personalized Movie Recommender System"
subtitle: "User-to-User Collaborative Filtering"
author: "Mark Hamer"
date: "2026-04-24"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: false
    number-sections: true
---

# Introduction

In the previous assignment, I used a **Global Baseline Estimate (GBE)** to predict movie ratings. The GBE is non-personalized, it uses the overall mean, a user bias, and a movie bias to make predictions, but it does not consider which specific users share similar taste. 

In this project, I build a **personalized recommender system** using **User-to-User Collaborative Filtering**. This algorithm identifies users who rate movies similarly to a target user (also known as neighbors), and uses their ratings to predict what the target user would rate an unseen movie.

**What the recommender outputs:**

- Predicted ratings for every unrated user movie pair
- A top-N ranked recommendation list per user
- RMSE evaluation via leave-one-out cross-validation, compared against the GBE baseline

# Step 1 — Load and Prepare the Data

```{r}
library(readxl)

ratings_data <- read_excel("MovieRatings.xlsx", sheet = "MovieRatings")

critic_names <- ratings_data$Critic

ratings_matrix <- as.matrix(ratings_data[, -1])
rownames(ratings_matrix) <- critic_names

ratings_matrix[ratings_matrix == "?"] <- NA
ratings_matrix <- apply(ratings_matrix, 2, as.numeric)
rownames(ratings_matrix) <- critic_names

overall_mean <- mean(ratings_matrix, na.rm = TRUE)

user_avg <- rowMeans(ratings_matrix, na.rm = TRUE)
user_bias <- user_avg - overall_mean

movie_avg <- colMeans(ratings_matrix, na.rm = TRUE)
movie_bias <- movie_avg - overall_mean

cat("Ratings Matrix:\n\n")
print(ratings_matrix)

cat("\nOverall mean:", round(overall_mean, 2), "\n")
```

# Step 2 — Center the Ratings

```{r}
centered_matrix <- ratings_matrix - user_avg

cat("Centered Ratings Matrix:\n\n")
print(round(centered_matrix, 2))
```

# Step 3 — Compute User Similarity

```{r}
n_users <- nrow(ratings_matrix)

sim_matrix <- matrix(NA, nrow = n_users, ncol = n_users)
rownames(sim_matrix) <- critic_names
colnames(sim_matrix) <- critic_names

for (i in 1:n_users) {
  for (j in 1:n_users) {
    if (i == j) {
      sim_matrix[i, j] <- 1  
      next
    }
    
    both_rated <- !is.na(centered_matrix[i, ]) & !is.na(centered_matrix[j, ])
    n_common <- sum(both_rated)
    
    if (n_common < 2) {
      sim_matrix[i, j] <- NA
      next
    }
    
    u_ratings <- centered_matrix[i, both_rated]
    v_ratings <- centered_matrix[j, both_rated]
    
    numerator <- sum(u_ratings * v_ratings)
    denominator <- sqrt(sum(u_ratings^2)) * sqrt(sum(v_ratings^2))
    
    if (denominator == 0) {
      sim_matrix[i, j] <- 0
    } else {
      sim_matrix[i, j] <- numerator / denominator
    }
  }
}

cat("User Similarity Matrix (Pearson Correlation):\n\n")
print(round(sim_matrix, 2))
```

# Step 4 — Generate Predictions

```{r}
predict_cf <- function(user, movie, k = 5) {
  
  if (!is.na(ratings_matrix[user, movie])) {
    return(ratings_matrix[user, movie])
  }
  
  sims <- sim_matrix[user, ]
  
  candidates <- which(!is.na(ratings_matrix[, movie]) & 
                       !is.na(sims) & 
                       sims > 0 & 
                       names(sims) != user)
  
  if (length(candidates) == 0) {
    prediction <- overall_mean + user_bias[user] + movie_bias[movie]
    return(prediction)
  }
  
  candidates <- candidates[order(sims[candidates], decreasing = TRUE)]
  if (length(candidates) > k) {
    candidates <- candidates[1:k]
  }
  neighbor_sims <- sims[candidates]
  neighbor_centered <- centered_matrix[candidates, movie]

  prediction <- user_avg[user] + 
    sum(neighbor_sims * neighbor_centered) / sum(abs(neighbor_sims))
  
  return(prediction)
}

prediction_matrix <- ratings_matrix

for (i in 1:nrow(ratings_matrix)) {
  for (j in 1:ncol(ratings_matrix)) {
    if (is.na(ratings_matrix[i, j])) {
      prediction_matrix[i, j] <- predict_cf(rownames(ratings_matrix)[i],
                                             colnames(ratings_matrix)[j])
    }
  }
}

cat("Predicted Ratings for All Missing Values:\n")
cat("(Original ratings kept as-is, predictions fill the gaps)\n\n")
print(round(prediction_matrix, 2))
```

# Step 5 — Top-N Recommendations Per User

```{r}
cat("Top Movie Recommendations Per User:\n")
cat("(Only showing movies the user has NOT rated, ranked by predicted rating)\n\n")

for (i in 1:nrow(ratings_matrix)) {
  user <- rownames(ratings_matrix)[i]
  
  unrated <- which(is.na(ratings_matrix[i, ]))
  
  if (length(unrated) == 0) {
    cat(user, ": Rated all movies — no recommendations needed\n\n")
    next
  }
  
preds <- setNames(prediction_matrix[i, unrated], colnames(ratings_matrix)[unrated])
  preds_sorted <- sort(preds, decreasing = TRUE)
  
  cat(user, ":\n")
  for (j in 1:length(preds_sorted)) {
    cat("  ", j, ". ", names(preds_sorted)[j], 
        " (predicted: ", round(preds_sorted[j], 2), ")\n", sep = "")
  }
  cat("\n")
}
```

# Step 6 — Evaluate with Leave-One-Out Cross-Validation

```{r}
cf_errors <- c()
gbe_errors <- c()

for (i in 1:nrow(ratings_matrix)) {
  for (j in 1:ncol(ratings_matrix)) {
    
    if (is.na(ratings_matrix[i, j])) next
    
    actual <- ratings_matrix[i, j]
    user <- rownames(ratings_matrix)[i]
    movie <- colnames(ratings_matrix)[j]
    
    temp_matrix <- ratings_matrix
    temp_matrix[i, j] <- NA
    
    if (sum(!is.na(temp_matrix[i, ])) < 2) next
    
    temp_user_avg <- rowMeans(temp_matrix, na.rm = TRUE)
    temp_centered <- temp_matrix - temp_user_avg
    
    temp_sim <- matrix(NA, nrow = nrow(temp_matrix), ncol = nrow(temp_matrix))
    rownames(temp_sim) <- rownames(temp_matrix)
    colnames(temp_sim) <- rownames(temp_matrix)
    
    for (a in 1:nrow(temp_matrix)) {
      for (b in 1:nrow(temp_matrix)) {
        if (a == b) { temp_sim[a, b] <- 1; next }
        both <- !is.na(temp_centered[a, ]) & !is.na(temp_centered[b, ])
        if (sum(both) < 2) next
        u_r <- temp_centered[a, both]
        v_r <- temp_centered[b, both]
        denom <- sqrt(sum(u_r^2)) * sqrt(sum(v_r^2))
        temp_sim[a, b] <- if (denom == 0) 0 else sum(u_r * v_r) / denom
      }
    }
    
    temp_overall <- mean(temp_matrix, na.rm = TRUE)
    gbe_pred <- temp_overall + 
      (temp_user_avg[user] - temp_overall) + 
      (mean(temp_matrix[, movie], na.rm = TRUE) - temp_overall)
    
    sims <- temp_sim[user, ]
    candidates <- which(!is.na(temp_matrix[, movie]) & 
                         !is.na(sims) & 
                         sims > 0 & 
                         names(sims) != user)
    
    if (length(candidates) == 0) {
      cf_pred <- gbe_pred  # Fall back to GBE
    } else {
      candidates <- candidates[order(sims[candidates], decreasing = TRUE)]
      if (length(candidates) > 5) candidates <- candidates[1:5]
      cf_pred <- temp_user_avg[user] + 
        sum(sims[candidates] * temp_centered[candidates, movie]) / 
        sum(abs(sims[candidates]))
    }
    
    if (!is.na(cf_pred) && !is.nan(cf_pred) && 
        !is.na(gbe_pred) && !is.nan(gbe_pred)) {
      cf_errors <- c(cf_errors, (actual - cf_pred)^2)
      gbe_errors <- c(gbe_errors, (actual - gbe_pred)^2)
    }
  }
}

cf_rmse <- sqrt(mean(cf_errors))
gbe_rmse <- sqrt(mean(gbe_errors))

cat("=== Model Evaluation (Leave-One-Out Cross-Validation) ===\n\n")
cat("Ratings evaluated:", length(cf_errors), "\n\n")
cat("Collaborative Filtering RMSE:", round(cf_rmse, 4), "\n")
cat("Global Baseline Estimate RMSE:", round(gbe_rmse, 4), "\n\n")

if (cf_rmse < gbe_rmse) {
  improvement <- round((1 - cf_rmse / gbe_rmse) * 100, 1)
  cat("Collaborative Filtering improves over GBE by", improvement, "%\n")
} else {
  cat("Global Baseline Estimate performed better on this dataset.\n")
  cat("This is not unusual with very sparse, small datasets where\n")
  cat("similarity estimates are unreliable.\n")
}
```

# Conclusion

## What Was Built

So for this project I implemented a **User-to-User Collaborative Filtering** recommender system using the same 16-critic, 6-movie survey dataset from the previous Global Baseline Estimate assignment. The system:

1. Centers each user's ratings by subtracting their personal average to remove scale differences
2. Computes Pearson correlation between all pairs of users (requiring at least 2 co-rated movies)
3. Predicts missing ratings using a weighted average of the most similar neighbors' centered ratings (up to k = 5 neighbors with positive similarity)
4. It'll fall back to the Global Baseline Estimate when no valid neighbors are available

## What It Outputs

- **Predicted ratings** for all 40 unrated user–movie pairs
- **Ranked recommendation lists** for each user, showing unseen movies ordered by predicted rating
- **RMSE scores** for both models via leave-one-out cross-validation

## Evaluation Results

| Global Baseline Estimate | 1.10 |
| Collaborative Filtering | 1.02 |

By using the collaborative filter we achieved a **7.7% improvement** over the non-personalized baseline. So, that demonstrates that by using user similarity it does improve prediction accuracy even on a small dataset.

## Limitations

- **Small and sparse data**: With only 16 users and 6 movies, similarity estimates are based on very few co-rated items, making them inherently noisy.
- **Out-of-range predictions**: The model can predict ratings outside the 1–5 scale (e.g., Dieudonne's predicted 6.0 for JungleBook). Clamping predictions to the valid range would address this.
- **Cold-start users**: Critics like Burton and Nathan with very few ratings cannot be effectively scored by collaborative filtering and we have to lean on the baseline fallback.